Interpolation is a process of finding new data points based on the range of a discrete set of known data points.
Interpolation has been well-developed in between 1674--1684
by Issac Newton's fundamental works, nowadays known as foundation of classical interpolation
theory~\cite{meijering2002chronology}.

The first time I found interpolation interesting was in 2016 when I observed a table of finite differences of cubes.
Back then, I was a first-year mechanical engineering undergraduate.
Due to my lack of mathematical knowledge, I started re-inventing interpolation
formulas myself, fueled by pure passion and a sense of mystery.
\textit{All the mathematical laws and relations exist from the very beginning; we only reveal and describe them},
I thought.
That mindset truly inspired me, and thus, my own mathematical journey began.

Consider finite differences of cubes $n^3$
\input{sections/figures/01_fig_finite_differences_cubes}
The problem of interpolation of polynomials is a classical problem in mathematics and has been widely studied in literature.
For instance, Concrete mathematics~\cite[p. 190]{graham1994concrete} gives interpolation of cubes by using
Newton's interpolation formula
\[
    n^3 = 6 \binom{n}{3} + 6 \binom{n}{2} + \binom{n}{1} + 0 \binom{n}{0}
\]
because
\begin{align*}
    f(x) = \Delta^{d} f(0) \binom{x}{d} +  \Delta^{d-1} f(0) \binom{x}{d-1} + \cdots + f(0) \binom{x}{0}
    = \sum_{r=0}^{d} \Delta^{d-r} f(0) \binom{x}{d-r}
\end{align*}
However, interpolation of cubes can be also done in a different way.
The key point that interpolation formula above iterates over the order $d$ of finite difference.
Alternatively, we can interpolate cubes $n^3$ as a sum of first order finite difference $\Delta$ as follows
\begin{align*}
    n^3 = \Delta 0^3 + \Delta 1^3 + \Delta 2^3 + \cdots + \Delta (n-1)^3 = \sum_{k=0}^{n-1} \Delta k^3
\end{align*}
We know that $\Delta^3 n^3 = 6$ is the constant for each $n$.
The second difference of cubes $\Delta^2 n^3$ is a linear relation in terms of third order finite difference
$\Delta^3 n^3$
\begin{align}
    \Delta^2 n^3 = (n+1) \Delta^3 n^3 = 6(n+1)
\end{align}
Finally, the first order finite difference $\Delta n^3$ is the following relation in terms of second
order finite difference
\begin{align*}
    \Delta n^3 = \Delta 0^3 + \Delta^2 0^3 + \Delta^2 1^3 + \cdots + \Delta^2 (n-1)^3 = 1 + \sum_{k=0}^{n-1} 6(k+1)
\end{align*}
Altering summation bounds yields
\begin{align*}
    \Delta n^3 = 1+6\cdot0+6\cdot1+6\cdot2+6\cdot3+\cdots+6\cdot n = 1 + 6 \sum_{k=0}^{n} k
\end{align*}
Therefore, we are able to express first order finite difference of cubes in form of sums as follows
\begin{align*}
    \Delta(0^3) &= 1+6 \cdot 0 \\
    \Delta(1^3) &= 1+6\cdot0+6\cdot1 \\
    \Delta(2^3) &= 1+6\cdot0+6\cdot1+6\cdot2 \\
    \Delta(3^3) &= 1+6\cdot0+6\cdot1+6\cdot2+6\cdot3
\end{align*}
Now it is time to assemble all the results above to get the polynomial $n^3$.
Having the relation in cubes $n^3 = \Delta 0^3 + \Delta 1^3 + \Delta 2^3 + \cdots + \Delta (n-1)^3$
we get
\begin{align*}
    n^3 &= [1+6\cdot0]+[1+6\cdot0+6\cdot1]+[1+6\cdot0+6\cdot1+6\cdot2] \nonumber \\
    &+ \cdots + [1+6\cdot0+6\cdot1+6\cdot2+\cdots+6\cdot(n-1)]
\end{align*}
By rearranging the terms of the equation above, we get summation in terms of $k (n-k)$
\begin{align}
    \label{eq:rearranged-cubes}
    n^3 = n &+ [(n-0) \cdot 6 \cdot 0] + [(n-1)\cdot6\cdot1] + [(n-2)\cdot6\cdot2] \nonumber \\
    &+ \cdots + [(n-k)\cdot 6 \cdot k] + \cdots + [1\cdot6\cdot(n-1)]
\end{align}
By applying compact sigma sum notation yields an identity for cubes $n^3$
\begin{align*}
    n^3 = n + \sum_{k=0}^{n-1} 6k(n-k)
\end{align*}
The term $n$ in the sum above can be moved under sigma notation, because there is exactly $n$ iterations, therefore
\begin{align*}
    n^3 = \sum_{k=0}^{n-1} 6k(n-k) + 1
\end{align*}
By inspecting the expression $6k(n-k) + 1$ we iterate under summation,
we can notice that it is symmetric over $k$, let be $T(n,k) = 6k(n-k) + 1$, then $T(n,k) = T(n,n-k)$.
This symmetry allows us to alter summation bounds again, so that
\begin{align*}
    n^3 = \sum_{k=1}^{n} 6k(n-k) + 1
\end{align*}
Curiously enough that although $\sum_{k=0}^{n-1} 6k(n-k) + 1$ and $\sum_{k=1}^{n} 6k(n-k) + 1$ both simplify to $n^3$,
they produce different closed forms.
Let be $P(n, q) = \sum_{k=0}^{q-1} 6k(n-k) + 1$ and $Q(n, q) = \sum_{k=1}^{q} 6k(n-k) + 1$, then
\begin{align*}
    P(n, q) =
    \begin{cases}
        q = 1: & 1 \\
        q = 2: & -4 + 6 n \\
        q = 3: & -27 + 18 n
    \end{cases}
\end{align*}
\begin{align*}
    Q(n, q) =
    \begin{cases}
        q = 1: & -5 + 6 n \\
        q = 2: & -28 + 18 n \\
        q = 3: & -81 + 36 n
    \end{cases}
\end{align*}
%Now let's take a breath and briefly summarize all the results we got so far.
%So, we have successfully interpolated the polynomial $n^3$ using discrete set of finite differences data points,
%by expressing its first order finite difference as $\Delta n^3 = 1+6\cdot0+6\cdot1+6\cdot2+6\cdot3+\cdots+6\cdot n$,
%so that $n^3$ is a sum of $\Delta n^3$.
%By rearranging the terms we got $n^3 = n + [(n-0) \cdot 6 \cdot 0] + [(n-1)\cdot6\cdot1] + [(n-2)\cdot6\cdot2]
%+ \cdots + [(n-k)\cdot 6 \cdot k] + \cdots + [1\cdot6\cdot(n-1)]$, leading to identities in terms of cubes
%\begin{align*}
%    n^3 = \sum_{k=0}^{n-1} 6k(n-k) + 1; \quad \quad n^3 = \sum_{k=1}^{n} 6k(n-k) + 1
%\end{align*}
%Indeed, summation bounds of above identities can be altered because $T(n,k) = 6k(n-k) + 1$
%is symmetric $T(n,k) = T(n,n-k)$.
